---
title: "Premier League Data Analysis"
author: "Teik Keat Tee"
date: "2023-12-12"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Introduction
The data of interest selected for this assignment was the Premier League data by Zaeem Nalla from Kaggle. 

Premier League (also known as 'Barclays Premier League' or 'English Premier League (EPL)') is the most pronounced and the highest level division of the English football league system. It was formed in February 20, 1992, and contested by 20 English clubs each season, following the English Football League's promotion and relegation system. Under this operation, we would observed changes of clubs for each season.

#### 1.1 Problem Statement
The Official Premier League data was mainly collected and processed by its exclusive statistics partner - 'Stats Perform'. According to the author of the Kaggle dataset, stats providers don't usually tend to release their data or analysed results for public access. This was true given that parties such as sports betting entities would be free from any form of cost constraints if it was the case. However, normal individuals like us might suffer from these constraints, when we only wish to improve our understanding about EPL teams and matches for personal interests. Furthermore, it would also be difficult for smaller-sized football club to form analysis about EPL matches and make better informed decisions towards improving their teams' performances in league matches.

#### 1.2 Objective
Given above problem statement, I am motivated to utilise the dataset extracted from Kaggle and perform some analysis on top of it. 

The general purpose of this assignment is to provide some insights about club performances within the season of 2006/07 to 2017/18. These insights can either be employed as a reference for future match outcome prediction, or as a top-level understanding for someone who's newly interested in EPL. The analysis covers a demonstration of top performing teams, key factors that affects the match outcomes, and interesting facts such as 'Offside Violation'.

#### 1.3 Data description
The raw dataset contains information of match results and team statistics for each season across 12 seasons, from season 2006/07 to season 2017/18. This dataset was originated from the Official Premier League website and scrapped, organised, and formatted by the author. This dataset comprises two tabular data formatted in csv files, namely results.csv and stats.csv. I aim to explore both data in a contemporaneous manner for the following analyses.

***

### 2. Data Preparation
Prior to the analyses, some steps should be taken care of. These include data extraction, data inspection, data preprocessing and importing dependencies.

#### 2.1 Import Dependencies
```{r import_dependencies, message=FALSE}
library(tidyverse) # [v2.0.0] For data wrangling
library(ggplot2) # [v3.4.4] For plotting
library(gridExtra) # [v2.3] For arranging plots
```

#### 2.2 Data Extraction
```{r data_extract, message=FALSE}
# Extracting the raw data from results.csv and stats.csv to a data frame
# Note that both csv files were stored in my github
results_df <- data.frame(read_csv("https://raw.githubusercontent.com/teikkeat80/Premier_League_Data_Analysis/main/premier_league_data/results.csv"))

stats_df <- data.frame(read_csv("https://raw.githubusercontent.com/teikkeat80/Premier_League_Data_Analysis/main/premier_league_data/stats.csv"))
```

#### 2.3 Data Inspection
Data inspection was always the first and foremost step to be conducted prior to performing analytical work. Through looking at the initial rows, structure and dimension of the data object, I am able to understand how the extracted data looks like from a top-level.

```{r results_initial_rows_inspection}
# Inspect initial rows of results.csv and stats.csv
head(results_df)
```

```{r stats_initial_rows_inspection}
head(stats_df)
```

```{r results_dimension_inspection}
# Inspect the dimensions of results.csv
print(paste0("[results table] rows: ", 
             dim(results_df)[1], 
             " columns: ",
             dim(results_df)[2]))
```

```{r stats_dimension_inspection}
# Inspect the dimensions of stats.csv
print(paste0("[stats table] rows: ", 
             dim(stats_df)[1], 
             " columns: ",
             dim(stats_df)[2]))
```

During the data extraction phase, as a result of dplyr's package - read_csv(), the column values were converted to appropriate data types. The dimensional inspection shows that results table consists of 4560 rows and 6 columns, while the stats table consists of 240 rows and 42 columns. Lastly, the initial rows of both tables also shows good condition of data, apart from some missing values identified in the stats table.

The results table shows no missing values for both categorical and numerical columns. No anomalies were detected for the categorical columns, as I may observed from the unique values and whitespace. For the numerical values, we can observe the maximum goals per match is 9 goals for home goals and 7 goals for away goals. Although these values were treated as outliers based on Interquartile Range, however it they were sensible values for our situation.

The stats table also shows no missing values for the categorical columns, however there were NAs appearing in some of the numerical columns, these include: "saves", "head_clearance", "total_through_ball", and "dispossessed" (20 NA values), and "backward_pass", "big_chance_missed" (80 NA values). These NA values were missing from the Premier League's website. However, these does not contains the columns being used in this analysis, where they would be removed in section 2.4. In terms of the anomaly values, no anomalies were detected in both categorical and numerical columns. Although there were some outliers, they were all sensible values.

Overall, the raw data were mostly cleaned up and presented in a good state. This is reasonable as it was obtained from Kaggle.

#### 2.4 Data Preprocessing
Despite the raw data is in a good quality, I aim to do some high-level pre-processing steps to make the flow of the analysis better. In particular, removing the irrelevant columns in stats_df (as all columns in results_df is relevant).
```{r stats_data_preprocess}
# Select only relevant columns from stats_df
# This is also a better practice to optimise code performance
stats_df <- stats_df %>% 
  select(team, season, goals, goals_conceded, wins, total_pass, total_scoring_att, total_offside)
```

Now the data is ready for the analysis!

**Note: Any further particular data transformation process will be considered during the analysis, if required.**

***

### 3. EPL Analysis
In this section I present the analysis for the performances and key statistics of all EPL football clubs which participated in seasons from 2006/07 to 2017/18.

#### 3.1 Top Performing Teams - *Analysis 1*
Firstly, it would be interesting to identify which clubs has won the most titles within these 12 seasons. Through observing the EPL title earners and top finishers, I could potentially find out the outstanding clubs of EPL.

EPL rank teams based on points and several tie-breaker rules, presented as follows in an orderly manner. For simplicity, I have only considered 3 of them, while it was usually rare to go any further.

1. Points
2. Goal Difference
3. Total Goals Scored

The points attributing system for each match is as follows:

1. Winning team gets 3 points
2. Losing team gets 0 points
3. Both teams gets 1 points in the occasion of a tie.

```{r analysis_1_dt1}
# Extract relevant columns from stats table
a1_goals <- stats_df %>% 
  select(team, season, goals, goals_conceded) %>% 
  # Calculate goal differences
  mutate(goal_diff = goals - goals_conceded)

# Calculate points obtained for each match from results table
# Alternatively, we can use the wins and losses from the stats table, but what if we only have the results table?
a1_pt <- results_df %>% 
  mutate(home_pt = if_else(result == "H", 3, if_else(result == "A", 0, 1)), 
         away_pt = if_else(result == "H", 0, if_else(result == "A", 3, 1)))

# Create a summarised data frame that calculate total points earned for each team in each season
a1_pt_sum <- a1_pt %>% 
  select(home_team, away_team, home_pt, away_pt, season) %>% 
  pivot_longer(cols = ends_with("pt"), names_to = "hoa", values_to = "pt") %>% 
  mutate(team = if_else(hoa == "home_pt", home_team, away_team)) %>% 
  select(-ends_with("_team"), -hoa) %>% 
  group_by(season, team) %>% 
  summarise(total_pt = sum(pt), .groups = 'drop') %>% 
  as.data.frame()

# Now that both tables has the same number of rows, I can proceed with joining both tables
a1_join <- inner_join(a1_goals, a1_pt_sum, by = c("season", "team")) %>% 
  select(-goals_conceded)

# Finds out the teams' position for each season and filter only top 3 teams
a1_pos <- a1_join %>% 
  group_by(season) %>% 
  arrange(desc(total_pt), desc(goal_diff), desc(goals)) %>% 
  mutate(position = row_number()) %>% 
  filter(position <= 3) %>% 
  ungroup()

# Summarise total champions and runner ups earned for each team
# The teams were ranked based on: Most Champion, 2nd Most Champion, ..., Most 1st runner up, ...
a1_sum <- a1_pos %>% 
  select(team, position) %>% 
  mutate(position = as.character(position), value = 1) %>% 
  pivot_wider(names_from = position, values_from = value, values_fn = sum, values_fill = 0) %>% 
  rename(champion = "1", first_runner_up = "2", sec_runner_up = "3") %>% 
  arrange(desc(champion), desc(first_runner_up), desc(sec_runner_up)) %>% 
  mutate(rank = row_number())
```

```{r analysis_1_plt1}
# Generate a stacked bar chart
a1_sum %>% 
  # Pivot to a longer shape to generate plot
  pivot_longer(cols = c(sec_runner_up, first_runner_up, champion), 
               names_to = "place", 
               values_to = "total") %>% 
  # Amend the place column to a factor for plotting
  mutate(place = recode_factor(
    factor(place, levels = c("sec_runner_up", "first_runner_up", "champion")), 
    sec_runner_up = "2nd Runner Up", 
    first_runner_up = "1st Runner Up", 
    champion = "Champion")) %>% 
  # Start plotting
  ggplot(aes(x = reorder(team, -rank), y = total, fill = place)) + 
  geom_bar(stat = "identity", width = 0.7) + 
  theme_bw() + 
  coord_flip() + 
  guides(fill = guide_legend(reverse = TRUE)) + 
  scale_fill_brewer(palette = "Reds") + 
  labs(title = "Title Winning Teams",
  subtitle = "Season: 2006/07 - 2017/18",
  x = "Team",
  y = "# of Title Wons",
  fill = "Place") + 
  theme(aspect.ratio = 1)
```

From above analysis, I have summarised 7 top performing teams in EPL across the 12 seasons, namely Manchester United, Chelsea, Manchester City, Leicester City, Liverpool, Arsenal, and Tottenham Hotspur. Manchester United was the best team among all, earning a total of 5 times EPL Champion titles, followed by Chelsea and Manchester City, earning 3 each. Nevertheless, these 3 teams came in the top 3 positions for most seasons. 

Moreover, 6 out of 7 teams here were generally perceived as the top teams in EPL. Therefore, it would be interesting to note that Leicester City has won a Champion title being an underdog during the 2015/16 season, where they started the season with odds of 1:5000 being the Champion. This was one of the greatest event in the EPL's history.

#### 3.2 Does passing or attacking leads to winning? - *Analysis 2*
A professional football team usually consists of 11 in-field players, with positions such as striker, goalkeeper, defender, and mid-fielder. The chemistry between all of these players is the key for the match outcome. Generally, the number of passes would determines how well the team cooperates in a football match. 

Apart from the collaboration of a team, offensive strategies were also important for winning a football match (as you can only win by scoring goals!). A team with good offensive tactics tends to create many scoring attempts in a match.

As a result, I aim to assess the relationship between passes and scoring attempts with winnings in this section.

```{r analysis_2_dt1}
# Extract relevant columns from stats table
a2_df <- stats_df %>% 
  select(wins, total_pass, total_scoring_att)
```

```{r analysis_2_plt1}
# Generate scatter plots to observe relationship between variables
a2_scat1 <- a2_df %>% 
  ggplot(aes(x = wins, y = total_pass)) + 
  geom_point() + 
  labs(title = "Pass vs. Wins",
       y = "# of Passes",
       x = "# of Games Won") + 
  theme_classic() + 
  theme(aspect.ratio = 1)

a2_scat2 <- a2_df %>% 
  ggplot(aes(x = wins, y = total_scoring_att)) + 
  geom_point() + 
  labs(title = "Scoring Attempt vs. Wins",
       y = "# of Scoring Attempts",
       x = "# of Games Won") + 
  theme_classic() + 
  theme(aspect.ratio = 1)

# Show the plot using grid.arrange (an alternative to plot_grid)
grid.arrange(a2_scat1, a2_scat2, nrow = 1, ncol = 2)
```

According the the scatter plots above, I observed a positive relationship in both plots. This explains that both passing and attempting to score matters! Nevertheless, it can also be explained as having a good team chemistry leads to creating scoring attempt and finally leads to winning a game. For brevity, I only aim to observe the relationship between these variables (instead of technical correlation effects). Ultimately, this analysis concludes that winning a football match hinges on a combination of good offensive strategy and seamless collaboration between all players.

#### 3.3 Offsides - *Analysis 3*
Offside was termed as a violation for the ball receiving player positioning themselves in front of the ball and the second last opponent, from the perspective of opponent's goal line. Whenever an offside call occurs, especially during a potential goal, it is always frustating to the players and their supporters. As such, it would be interesting to understand the offside dynamics thorough out the EPL's 12 seasons.

```{r analysis_3_dt1}
# Extract columns of interest from stats table
a3_df <- stats_df %>% 
  select(season, total_offside)

# To observe the dynamics about how did the number of offside calls evolve until recent years, it is sensible to create a line graph.
# Therefore, I transform the data to a time series with continuous variables such as date variables.
a3_ts <- a3_df %>% 
  group_by(season) %>% 
  summarise(sum_offside = sum(total_offside)) %>% 
  # I will be using the beginning of the year of season start as the date variable
  mutate(season = as.Date(
      paste0(sub(pattern = "(\\d+)-.*", replacement = "\\1", x = season), 
             "-01-01")))
```

```{r analysis_3_plt1}
# Generate line plot
a3_ts %>% 
  ggplot(aes(x = season, y = sum_offside)) + 
  geom_point() + 
  geom_line() + 
  # Using the anonymous function and %m+% (from tidyverse's lubridate package) to create date labels like year/year+1 (e.g. 06/07) for display
  scale_x_date(labels = function (x) paste0(format(x, "%y"), "/", format(x %m+% years(1), "%y"))) + 
  theme_minimal() + 
  labs(title = "Total Offside Calls over Seasons",
       subtitle = "Season: 2006/07 - 2017/18",
       x = "Season",
       y = "# of Offside Calls") + 
  theme(aspect.ratio = 1)
```

The line graph above shows a downward trend in offside violations throughout the years (with the exception of it going up again in the 2017/18 season). This could be attributed to players improved in terms of falling into offside positions from time to time. It could also be explained as more and more players managed to get through the violation under the eyes of referees, which could be due to a quality drop in referees or managers' tricks. This indicates that solely relying on referees could be insufficient, therefore, advanced technologies such as the VAR were developed and employed in more recent days.

#### 3.4 Home wins vs. Away wins - *Analysis 4*
Generally, home teams were perceived as the side with higher winning odds, especially when both opposing teams were on even level. This is due to the home side's advantage on familiarity with the field, larger supporting crowd, and players' belonging feelings. In this section, I aim to provide a comparative analysis on home and away winnings.

```{r analysis_4_dt1}
# Create a copy of the results table for this section
a4_df <- results_df

# Summarise home and away wins for each teams
a4_sum <- a4_df %>% 
  filter(!result == "D") %>% 
  mutate(win_team = if_else(result == "H", home_team, away_team)) %>% 
  group_by(win_team) %>% 
  summarise(
    home_wins = sum(result == "H"),
    away_wins = sum(result == "A")
  )
```

```{r, analysis_4_plt1, fig.height=4, fig.width=4}
# Generate a pie chart for proportionally comparing all home and away wins throughout the years.
a4_sum %>% 
  select(home_wins, away_wins) %>% 
  pivot_longer(cols = c(away_wins, home_wins), 
               names_to = "win_type", 
               values_to = "win_total") %>% 
  group_by(win_type) %>% 
  summarise(win_total = sum(win_total)) %>% 
  mutate(win_type = str_to_title(gsub("_wins", "", win_type))) %>% 
  ggplot(aes(x = "", y = win_total, fill = win_type)) + 
  geom_bar(stat = "identity", width = 1, color = "white") + 
  # Labeling the percentages
  geom_text(aes(label = win_total, fontface = "bold"), 
            size = 5, 
            color = "white", 
            position = position_stack(vjust = 0.5)) + 
  coord_polar("y") + 
  theme_void() + 
  theme(aspect.ratio = 1) + 
  labs(
    title = "Home's side victory vs. Away's side victory",
    fill = "Home/Away"
  ) + 
  guides(fill = guide_legend(reverse = TRUE)) + 
  scale_fill_brewer(palette = "Set1")
```

According to the pie chart above, I observed that generally EPL teams tends to win the match when they are playing at their home court. Out of all non-tying matches, home side's victory were 63% more than away side's victory. This piece of information would be important for football fans to form expectations before a match began, and for football teams to plan a better strategy for which side their team is playing on in future games.

#### 3.5 Is it easy to score a goal? - *Analysis 5*
Other sports enthusiasts may argue that a football match can be dull to watch, given its long duration with relatively few exciting scoring moments. Nevertheless, the ability to defend from conceding a goal was also a key factor for the team to win the match. Furthermore, the goal differences (goals scored - goals conceded) was sometimes a major tie breaker for a team to win the championship. Therefore, let's try to analyse the ability of a team to defend its goal.

```{r analysis_5_dt1}
# Extract relevant data from the stats table
a5_df <- stats_df %>% 
  select(team, goals_conceded, season)
```

```{r analysis_5_plt1}
# Box plots might be a good way to showcase the goals conceded statistics
# Moreover, I will also aim to emphasise on the 'Big 4' teams in EPL, which include Manchester United, Arsenal, Chelsea, and Liverpool.
a5_df %>% 
  # Showing season label in a better presentation
  mutate(season = sub("\\d{2}(\\d{2})-\\d{2}(\\d{2})", "\\1/\\2", season)) %>% 
  ggplot(aes(x = season, y = goals_conceded, fill = season)) + 
  geom_boxplot(fill = NA, color = "grey") + 
  theme_classic() + 
  geom_text(aes(
    label = if_else(
      team %in% c("Manchester United", "Arsenal", "Chelsea", "Liverpool"), 
      substr(team, 1, 1), ""), 
    color = team, 
    fontface = "bold")) + 
  scale_color_manual(values = c("Manchester United" = "red", 
                                "Arsenal" = "black", 
                                "Chelsea" = "blue", 
                                "Liverpool" = "orange")) + 
  theme(legend.position = "None") + 
  labs(
    title = "Goals Conceded Box Plots",
    subtitle = "Season: 2006/07 - 2017/18",
    x = "Season",
    y = "# of Goals Conceded"
  )
```

From above box plots, I observed that the median goals conceded fluctuates around 40-60 goals per season. Given that a team would usually plays 38 matches in an EPL season, we should, on average, seeing one goal per match for each season. The 07/08 season was by far the most interesting season that having a team losing the minimum goals and another team losing the maximum goals, across all 12 seasons.

Apart from that, the illustration of 'Big 4' EPL teams in the box plot shows that these teams tend to concede lesser goals in comparison to other teams, with an exception of 15/16's Chelsea. This implies 'great teams comes with great defense'. It is also worthy to note that there were two outlying values in seasons 13/14 and 14/15 respectively. These were 13/14's Fulham and 14/15's Queens Park Rangers. Undoubtly, they came in last 3 positions when the season ends and relegated.

Overall, this analysis shows us the importance of defensive strategy for a team to achieve success in EPL. While it might be discouraging that some matches have little to none scoring moments, the defensive play in football would still be often quite exciting.

***

### 4. Conclusion and remarks
A better knowledge towards how a football game/league works would make a match (or the whole season!) even more exciting to watch. As shown in above analyses, factors that make a football team strong consists of not only if they have many good players, but also their offensive and defensive strategies, team chemistry, playing ground, familiarity with rules, and many more. As such, club managers should always consider holistically. Nevertheless, these factors also contributed to making EPL or general football games more interesting.

### Session Info
```{r}
sessionInfo()
```